---
title: "Trajectory Mapping Results"
subtitle: "analysis x"
author: "Cecilia Valenzuela"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  beamer_presentation:
    theme: "Pittsburgh"
    colortheme: "dove"
    fonttheme: "structurebold"
    highlight: pygments
classoption: "aspectratio=169"
fontsize: 8pt
tables: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(knitr.kable.NA = '')
```

```{r libraries}
# Load libraries ---------------------------------------------------------------
library(argparse)
library(yaml)
library(tidyverse)
library(lubridate)
library(scales)
library(ggpubr)
library(ggsci)
library(ggridges)
library(patchwork)
library(hrbrthemes)
library(circlize)
library(chorddiag) 
library(ggforce)
library(ComplexHeatmap)
library(maps)
library(sf)
library(wpp2019)
library(gridExtra)
library(grid)
# Source files -----------------------------------------------------------------
source("./scripts/trajProcessing.R")
source("./scripts/utils.R")
```

```{r parser}
# Parser -----------------------------------------------------------------------
parser <- argparse::ArgumentParser()
parser$add_argument("--input", type = "character", help="trajectory file")
parser$add_argument("--burnin", type = "double", help = "Burning fraction for trajectory files")
parser$add_argument("--metadata", type = "character", help = "Alignment sequences metadata")
parser$add_argument("--demes", nargs = "+", help = "Demes configuration")
parser$add_argument("--n", type = "integer", help = "Number of trajectories to analyze")
parser$add_argument("--output_figure", type = "character", help = "Output path for the figures")
parser$add_argument("--output_table", type = "character", help = "Output path for the tables")
args <- parser$parse_args()
```

```{r debugging}
# Debugging
args$burnin = 0.1
args$demes = read.csv("./files/demes.csv", stringsAsFactors = FALSE)
args$input = "./results/europeC/201106_europe5.TL.traj"
args$metadata =  "./results/europeC/final_metadata.tsv"
args$n = 50
demes = args$demes
demes$exclude_country[5] = "France,Germany,Italy,Spain"
demes$deme[3] = "China"
demes$division[3] = NA
demes <- demes %>% arrange(deme)
```




```{r plotopts}
# Plot options -----------------------------------------------------------------------
set_plotopts()
dcolors <- pal_npg("nrc")(10)[c(1:5,9)]
dcolors <- dcolors[1:nrow(demes)]
names(dcolors) <- sort(demes$deme)
dark_dcolors <- rgb(t(col2rgb(dcolors)/2), maxColorValue=255)
names(dark_dcolors) <- names(dcolors)
ecolors <- c("#1e3d59", "#f5f0e1", "#ff6e40", "#ffc13b")
names(ecolors) <- c("B", "M", "IM", "OM")
```


## Deme configuration
\tiny
```{r demeconfig}
# Load demes configuration and sequence metadata -------------------------------
# demes <- data.frame(deme = NA, region = NA, country = NA, division = NA, 
#                     exclude_country = NA, exclude_division = NA,
#                     min_date = NA, max_date = NA, seq_per_deme = NA) %>%
#   bind_rows(bind_rows(lapply(yaml::yaml.load(string = paste(args$demes, collapse = " ")),
#                           data.frame, stringsAsFactors = FALSE), .id = "deme")) %>%
#   group_by(deme) %>% 
#   mutate(exclude_country = ifelse(is.na(exclude_country), NA, paste0(exclude_country, collapse = ","))) %>%
#   distinct() %>%
#   filter_all(any_vars(!is.na(.))) %>%
#   mutate(deme = str_to_title(deme)) %>%
#   mutate(deme = ifelse(deme %in% c("Othereuropean", "OtherEuropean"), "Other European", deme))

china_deme <- demes$deme[demes$region == "Asia"]
if (china_deme == "Hubei-China") china_deme = "China"
```

```{r map, fig.height=2, fig.align="center"}
# https://www.r-bloggers.com/2019/04/zooming-in-on-maps-with-sf-and-ggplot2/
worldmap <- st_as_sf(map("world", plot = FALSE, fill = TRUE)) %>%
  mutate(country = ifelse(ID == "UK", "United Kingdom", as.character(ID)),
         continent = countrycode(sourcevar = country,
                             origin = "country.name",
                             destination = "continent")) %>%
  #filter(country %in% demes$country | continent == "Europe") %>%
  mutate(deme = case_when(
    country %in% demes$country ~ country,
    continent == "Europe" ~ "OtherEuropean"))

zoom_to <- c(20, 50) 
zoom_level <- 2.5
target_crs <- sprintf('+proj=aeqd +lon_0=%f +lat_0=%f',
                      zoom_to[1], zoom_to[2])
C <- 40075016.686   # ~ circumference of Earth in meters
x_span <- C / 2^zoom_level
y_span <- C / 2^(zoom_level+1)

zoom_to_xy <- st_transform(st_sfc(st_point(zoom_to), crs = 4326),
                           crs = target_crs)
disp_window <- st_sfc(
    st_point(st_coordinates(zoom_to_xy - c(x_span/2, y_span/2))),
    st_point(st_coordinates(zoom_to_xy + c(x_span *1.2, y_span *1.2))),
    crs = target_crs
)
ggplot() + geom_sf(data = worldmap, aes(fill = deme), alpha = 0.8, size = 0) +
    scale_x_continuous(breaks = seq(-100, 200, by = 10)) +
    scale_y_continuous(breaks = seq(0, 100, by = 10)) +
    #geom_sf(data = zoom_to_xy, color = 'red') +
    coord_sf(xlim = st_coordinates(disp_window)[,'X'],
             ylim = st_coordinates(disp_window)[,'Y'],
             crs = target_crs, expand = FALSE) +
  scale_fill_manual(values = dcolors, na.value = "grey90") +
  theme_void() +
  theme(panel.grid.major = element_line(colour = "grey", size = 0.1),
        legend.position = "none")
        # panel.ontop = TRUE,
        # panel.background = element_blank()) 

knitr::kable(demes %>% select(-c("demeID", "exclude_division", "prob")), booktabs = TRUE)
```


## ECDC Case count data

```{r ecdc-data, results='hide'}
# Case data information from ECDC ----------------------------------------------
# Most recent sample value
metadata <- read_tsv(args$metadata)
MRS <- max(metadata$date)
mrs <- ymd_hms(paste(MRS, '23:59:00 UTC'))

case_data <- get_cases(demes, from = ymd("2019-09-01"), to = MRS + 10) %>%
  arrange(date) %>%
  group_by(deme) %>%
  rename(value = cases) %>%
  mutate(cumvalue = cumsum(value)) %>%
  group_by(deme, date) %>%
  mutate(cumvalue = max(cumvalue)) %>%
  select(deme, date, cumvalue, value)  %>%
  distinct_at(vars(deme, date, cumvalue), .keep_all = TRUE)

# Population data
data(pop)
population <- pop %>% filter(name %in% demes$deme) %>% select(name, "2020")
pop_eu_demes <- population %>% filter(name != "China") %>% summarise(s = sum(`2020`)) %>% pull(s)
pop_eu_total <- pop %>% filter(name == "Europe") %>% pull("2020")
population <- population %>%
  tibble::add_row(name = "OtherEuropean", `2020` = pop_eu_total - pop_eu_demes) %>% rename(deme = name, pop = 2)

case_data_pop <- case_data %>% left_join(population, by="deme") %>% mutate(ECDCcas100 = cumcases * 1e5 / (pop * 1e3)) 
```

::: {.columns}

:::: {.column width="40%"}
\small
```{r ecdc-table, echo=FALSE}
knitr::kable(case_data_pop %>% ungroup %>% filter(date == MRS + 10) %>% select(deme, cumcases, pop, cas100) %>% mutate_if(is.numeric, function (x) round(x,1)), booktabs = TRUE, caption = "Total number of cases reported to ECDC 18th March")
```
::::

:::: {.column width="60%"}
```{r ecdc-plot, fig.height = 3, fig.width=4, fig.cap="ECDC case counts for each deme from the beginning of the pandemic to March 18"}
ggplot(case_data) +
  geom_line(aes(x = date, y = cumcases, color = deme)) +
   scale_color_manual(values = dcolors, name = "") +
   scale_x_date(limits = c(ymd("2019-12-30"), ymd("2020-03-18")), date_breaks = "1 month", date_labels = "%b %d") +
    scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
    ylab("Case count") +
    theme(axis.title.x=element_blank(),
          legend.position = c(0.2, 0.7),
          legend.box="horizontal",
          panel.grid.major.x =  element_line(colour = "grey80", size = 0.3),
          panel.grid.minor.x =  element_line(colour = "grey80", size = 0.2))
```

::::

:::

## Epidemic trajectory data

From the Stochastic Trajectory Mapping analysis, we obtain one epidemic trajectory per set of parameters + typed node tree.

The processing of the trajectory data includes the generation of two different datasets:

- `states`: We have the total number of inferred cases by trajectory, deme and time.
- `events`: We have each event that happened in a epidemic trajectory, with its type (origin, birth, death or migration), the source/destination deme and time.


```{r traj-data, warnings = FALSE, message = FALSE}
# Load trajectory data ---------------------------------------------------------
df <- loadTrajectories(args$input, burninFrac = args$burnin, subsample = args$n)

# Data wrangling ---------------------------------------------------------------
states <- df$states %>%
  mutate(type = factor(type,
                       levels = 0:(nrow(demes) - 1),
                       #labels = sort(demes$deme)),
                       labels = c("France", "Germany","China","Italy","OtherEuropean","Spain")),
         date_model = date(date_decimal(decimal_date(ymd(MRS)) - age))) %>%
  mutate(date = date_model + 10)

# events <- df$events %>%
#   mutate(src = factor(src,
#                       levels = 0:(nrow(demes) - 1),
#                       #labels = sort(demes$deme)),
#                       labels = c("France", "Germany","China","Italy","OtherEuropean","Spain")),
#          dest = ifelse(is.na(dest), as.character(src),
#                        as.character(factor(dest,
#                                            levels = 0:(nrow(demes) - 1),
#                                            #labels = sort(demes$deme),
#                                            labels = c("France" , "Germany","China","Italy","OtherEuropean","Spain")))),
#          date_model = date(date_decimal(decimal_date(ymd(MRS)) - age))) 


# Do this in a function, useful also for Claire
events <- df$events %>%
  mutate(src = factor(src,
                      levels = 0:(nrow(demes) - 1),
                      #labels = sort(demes$deme)),
                      labels = c("France", "Germany","China","Italy","OtherEuropean","Spain")),
         dest = (factor(dest,
                           levels = 0:(nrow(demes) - 1),
                           #labels = sort(demes$deme),
                           labels = c("France" , "Germany","China","Italy","OtherEuropean","Spain"))))
events[events$event == "O", "src"] = "China"
events[events$event == "O", "mult"] =  1

df_traj0 <- events %>%
  mutate(date = date(date_decimal(decimal_date(mrs) - age))) %>%
  select(-time, -age) %>%
  group_by_at(vars(-mult)) %>%
  summarise(N = sum(mult), .groups = "drop") %>%
  mutate(srctmp = src, destmp = dest) %>%
  pivot_longer(src:dest, names_to = "role", values_to = "deme") %>%
  mutate(event = ifelse(event == "M" & role == "src", "OM",
                        ifelse(event == "M" & role == "dest", "IM", event)),
         partner = ifelse(event == "OM", as.character(destmp),
                        ifelse(event == "IM", as.character(srctmp), NA))) %>% 
  select(-c("role", "srctmp", "destmp")) %>%
  filter(!is.na(deme)) %>%
  group_by_at(vars(-N)) %>%
  summarise(value = sum(N), .groups = "drop") %>%
  group_by(traj, deme, partner, event) %>%
  arrange(traj, date, deme) %>%
  mutate(cumvalue = cumsum(value)) %>%
  rename(var = event) 

mig_cum = df_traj0 %>%
  filter(var %in% c("IM", "OM")) %>%
  group_by(traj, var, date, deme) %>%
  summarise(value = sum(value), .groups = "drop") %>%
  group_by(traj, var, deme) %>%
  arrange(traj, date, deme) %>%
  mutate(cumvalue = cumsum(value),
         partner = NA) 
            
inout_pop = df_traj0 %>% 
  mutate(var = ifelse(var %in% c("O", "B", "IM"), "in_pop", "out_pop")) %>%
  group_by(traj, deme, var, date) %>%
  summarise(value = sum(value), .groups = "drop_last") %>%
  arrange(traj, date, deme) %>%
  mutate(cumvalue = cumsum(value)) 

active_pop = inout_pop %>%
  select(-cumvalue) %>%
  pivot_wider(names_from = var, values_from = value, values_fill = list(value = 0)) %>%
  mutate(active_pop = in_pop - out_pop) %>%
  pivot_longer(active_pop, names_to = "var", values_to = "value") %>%
  group_by(traj, deme, var) %>%
  arrange(traj, date, deme) %>%
  mutate(cumvalue = cumsum(value)) %>%
  select(-c("in_pop", "out_pop"))

df_traj <- bind_rows(df_traj0, mig_cum) %>%
  bind_rows(inout_pop) %>%
  bind_rows(active_pop) 

df_traj_cases <- df_traj %>%
  mutate(traj = as.character(traj)) %>%
  bind_rows(case_data %>%
            filter(cumvalue != 0) %>%
            mutate(var = "ECDC"))
                  # traj = "ecdc"))
```

```{r table-states}
knitr::kable(head(states, 5), booktabs = TRUE, caption = "States dataset")
```

## Epidemic trajectory data

```{r table-events}
knitr::kable(head(events, 5), booktabs = TRUE, caption = "Events dataset")
```

\normalsize
To have a feasible time of analysis of the epidemic trajectories we take a random subsample of 500 trajectories.

To facilitate visualization and summarise the results, we take a grid time of 1 day and summarise the number of events that day as the sum of the events in the corresponding time interval; and the number of inferred cases as the maximum of the interval.


```{r traj-process, warning=FALSE}
# # Date grids
# max_age <- max(states$age)
# min_date <- date_decimal(decimal_date(mrs) - max_age)
# #ages_5day <- rev(decimal_date(mrs) - decimal_date(seq(mrs,min_date, by = '-5 days')))
# ages_1day <- rev(decimal_date(mrs) - decimal_date(seq(mrs,min_date, by = '-1 days')))
# 
# # Grid trajectories, time grid of 5 days taking maximum value
# gt <- gridTrajectoriesByAge(states, ages_1day) %>%
#   replace_na(list(N = 0)) %>%
#   mutate(date_model = date(date_decimal(decimal_date(mrs) - age)),
#          date = date_model + 10)
# 
# # Summary statistics
# gt_summary <- gt %>%
#   group_by(date, type) %>%
#   summarize(Imean = mean(N),
#             Imedian = median(N),
#             Ilow    = quantile(N, 0.025), #95% CI, compute HPD??
#             Ihigh   = quantile(N, 0.975),
#             .groups = "drop_last") %>%
#   ungroup()
# 
# # Grid events, time grid of 1 day, taking the sum of events
# ge <- gridEventsByAge(events, ages_1day) %>%
#   mutate(date_model = date(date_decimal(decimal_date(mrs) - age)),
#          date = date_model + 10,
#          dest = ifelse(is.na(dest), as.character(src), as.character(dest))) %>%
#   ungroup %>%
#   filter(N != 0) 

```


<!-- ## Inferred case counts -->

```{r}
# 1. Trajectories - Total case counts
# 1.1 Median and 95% CCI of Total cases on the full time period. Fraction median
#    and 95% CCI (cases ecdc/inferred cases) wrt. ECDC reported cases

reported <- gt_summary %>%
  rename(deme = type) %>%
  filter(date == MRS + 10) %>%
  left_join(case_data_pop , by = c("deme", "date")) %>%
  mutate(fc = cumcases/Imedian,
         lfc = cumcases/Ihigh,
         hfc = cumcases/Ilow,
         rc = Imedian/cumcases,
         lrc = Ilow/cumcases,
         hrc = Ihigh/cumcases,
         mcas100 = Imedian * 100 / pop,
         lcas100 = Ilow * 100 / pop,
         hcas100 = Ihigh * 100 / pop) %>%
  #select(deme, Imedian, Ilow, Ihigh, rc, lrc, hrc, cumcases, pop, cas100) %>%
  mutate_if(is.numeric, function(x) round(x,2))

# r_reported <- reported %>%
#   mutate(Country = type,
#          `Cases ECDC` = cumcases,
#          `Median Cases (95% CCI)` = paste0(round(Imedian), " (", round(Ilow), "-", round(Ihigh), ")"),
#          `Median Reporting Rate (95% CCI)` = paste0(round(reportrate, 2), " (", round(hreportrate, 2), "-", round(lreportrate, 2), ")")) %>%
#   select(-one_of(colnames(reported)))
#
# write_tsv(r_reported, paste0(args$output_table, "01.tsv"))
```

```{r trajs}
# 1.2 All trajectories single deme in log scale with ECDC case count data
trajs <- lapply(demes$deme, function(deme) {
  gt %>%
    # 1. Prepare data
    filter(type == deme) %>%
    mutate(traj = as.character(traj)) %>%
    bind_rows(case_data %>%
                rename(type = deme,
                       N = cumcases) %>%
                filter(type == deme) %>%
                mutate(traj = "ecdc") %>%
                select(type, traj, date, N) %>%
                distinct()) %>%
    # 2. Plot
    ggplot() +
    geom_line(aes(date, N, group = traj,
                  color = traj == "ecdc",
                  alpha = traj == "ecdc",
                  linetype = traj == "ecdc",
                  size = traj == "ecdc")) +
    scale_color_manual(name = "",
                       labels = c(paste0(deme, " inferred\n population trajectories"),
                                  paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
                                         " ECDC total case count")),
                       values = c(dcolors[[deme]], dark_dcolors[[deme]])) +
    scale_alpha_manual(name = "", values = c(0.2, 1), guide = FALSE) +
    scale_linetype_manual(name = "", values = c(1, 2),
                          labels = c(paste0(deme, " inferred\n population trajectories"),
                                     paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
                                            " ECDC total case count"))) +
    scale_size_manual(name = "", values = c(0.5, 1),
                      labels = c(paste0(deme, " inferred\n population trajectories"),
                                 paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
                                        " ECDC total case count"))) +
    scale_x_date(limits = c(ymd("2019-11-01"), ymd(MRS + 10)), date_breaks = "2 weeks", 
                 date_labels = "%b %d") +
    scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
    ylab("Population size") +
    theme(legend.position = c(0.4, 0.9),
          axis.title.x=element_blank())
  })

names(trajs) <- demes$deme
# ggexport(ggarrange(plotlist = trajs, labels = chartr("123456789", "ABCDEFGHI", 1:length(trajs)), nrow = 3, ncol = 2),
#          filename = paste0(args$output_figure, "02.png"),
#          width = 2300, height = 3000, res = 300)
```

```{r}
# Active cases
d = "France"
df_traj_cases %>%
    # 1. Prepare data
    filter(deme == d, var %in% c("active_pop", "in_pop", "ECDC"), is.na(partner)) %>%
    # 2. Plot
    ggplot() +
    geom_line(aes(date, cumvalue, group = traj,
                  color = var)) +
                  # color = traj == "ecdc",
                  # alpha = traj == "ecdc",
                  # linetype = traj == "ecdc",
                  # size = traj == "ecdc")) +
    # scale_color_manual(name = "",
    #                    labels = c(paste0(deme, " inferred\n population trajectories"),
    #                               paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
    #                                      " ECDC total case count")),
    #                    values = c(dcolors[[deme]], dark_dcolors[[deme]])) +
    # scale_alpha_manual(name = "", values = c(0.2, 1), guide = FALSE) +
    # scale_linetype_manual(name = "", values = c(1, 2),
    #                       labels = c(paste0(deme, " inferred\n population trajectories"),
    #                                  paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
    #                                         " ECDC total case count"))) +
    # scale_size_manual(name = "", values = c(0.5, 1),
    #                   labels = c(paste0(deme, " inferred\n population trajectories"),
    #                              paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
    #                                     " ECDC total case count"))) +
    scale_x_date(limits = c(ymd("2019-11-01"), ymd(MRS + 10)), date_breaks = "2 weeks", 
                 date_labels = "%b %d") +
    scale_y_log10(expand = c(0, 0),labels = trans_format("log10", math_format(10^.x))) +
    ylab("Population size") +
    theme(legend.position = c(0.4, 0.9),
          axis.title.x=element_blank())
```

\tiny
```{r trajs-plotv2, results='asis', fig.height = 4}
for (d in demes$deme) {
  cat("\n## Inferred case counts - ", d, "\n")
  grid.arrange(trajs[[d]], tableGrob(reported %>% filter(deme == d) %>% t, theme = ttheme_minimal()), ncol = 2, widths = c(2,1))
  cat("\n")
}
```
```{r}
gt_v2 <- ge %>%
  ungroup %>%
  filter(event %in% c("B", "M")) %>%
  group_by(traj, dest, date) %>%
  arrange(date) %>%
  summarise(Nt = sum(N), .groups = 'drop_last') %>%
  mutate(cumNt = cumsum(Nt)) %>%
  rename(type = dest)

gt_summary_v2 <- gt_v2 %>%
  group_by(date, type) %>%
  summarize(Imean = mean(cumNt),
            Imedian = median(cumNt),
            Ilow    = quantile(cumNt, 0.025), #HPD??
            Ihigh   = quantile(cumNt, 0.975),
            Dmedian = median(Nt),
            Dlow    = quantile(Nt, 0.025), #HPD??
            Dhigh   = quantile(Nt, 0.975),
            .groups = "drop_last") %>%
  ungroup()

```

```{r trajs-infov2}
# 1. Trajectories - Total case counts
# 1.1 Median and 95% CCI of Total cases on the full time period. Fraction median
#    and 95% CCI (cases ecdc/inferred cases) wrt. ECDC reported cases

reported <- gt_summary_v2 %>%
  rename(deme = type) %>%
  filter(date == MRS + 10) %>%
  left_join(case_data_pop , by = c("deme", "date")) %>%
  mutate(fc = cumcases/Imedian,
         lfc = cumcases/Ihigh,
         hfc = cumcases/Ilow,
         rc = Imedian/cumcases,
         lrc = Ilow/cumcases,
         hrc = Ihigh/cumcases,
         mcas100 = Imedian * 100 / pop,
         lcas100 = Ilow * 100 / pop,
         hcas100 = Ihigh * 100 / pop) %>%
  #select(deme, Imedian, Ilow, Ihigh, rc, lrc, hrc, cumcases, pop, cas100) %>%
  mutate_if(is.numeric, function(x) round(x,2))

# r_reported <- reported %>%
#   mutate(Country = type,
#          `Cases ECDC` = cumcases,
#          `Median Cases (95% CCI)` = paste0(round(Imedian), " (", round(Ilow), "-", round(Ihigh), ")"),
#          `Median Reporting Rate (95% CCI)` = paste0(round(reportrate, 2), " (", round(hreportrate, 2), "-", round(lreportrate, 2), ")")) %>%
#   select(-one_of(colnames(reported)))
#
# write_tsv(r_reported, paste0(args$output_table, "01.tsv"))
```

```{r trajsv2}
# 1.2 All trajectories single deme in log scale with ECDC case count data
trajs <- lapply(demes$deme, function(deme) {
  gt_v2 %>%
    # 1. Prepare data
    filter(type == deme) %>%
    mutate(traj = as.character(traj)) %>%
    bind_rows(case_data %>%
                rename(type = deme,
                       cumNt = cumcases) %>%
                filter(type == deme) %>%
                mutate(traj = "ecdc") %>%
                select(type, traj, date, cumNt) %>%
                distinct()) %>%
    # 2. Plot
    ggplot() +
    geom_line(aes(date, cumNt, group = traj,
                  color = traj == "ecdc",
                  alpha = traj == "ecdc",
                  linetype = traj == "ecdc",
                  size = traj == "ecdc")) +
    scale_color_manual(name = "",
                       labels = c(paste0(deme, " inferred\n population trajectories"),
                                  paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
                                         " ECDC total case count")),
                       values = c(dcolors[[deme]], dark_dcolors[[deme]])) +
    scale_alpha_manual(name = "", values = c(0.2, 1), guide = FALSE) +
    scale_linetype_manual(name = "", values = c(1, 2),
                          labels = c(paste0(deme, " inferred\n population trajectories"),
                                     paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
                                            " ECDC total case count"))) +
    scale_size_manual(name = "", values = c(0.5, 1),
                      labels = c(paste0(deme, " inferred\n population trajectories"),
                                 paste0(ifelse(deme == china_deme, "China", as.character(deme)), 
                                        " ECDC total case count"))) +
    scale_x_date(limits = c(ymd("2019-11-01"), ymd(MRS + 10)), date_breaks = "2 weeks", 
                 date_labels = "%b %d") +
    scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
    ylab("Population size") +
    theme(legend.position = c(0.3, 0.9),
          axis.title.x=element_blank())
  })

names(trajs) <- demes$deme
# ggexport(ggarrange(plotlist = trajs, labels = chartr("123456789", "ABCDEFGHI", 1:length(trajs)), nrow = 3, ncol = 2),
#          filename = paste0(args$output_figure, "02.png"),
#          width = 2300, height = 3000, res = 300)
```


\tiny
```{r trajs-plot, results='asis', fig.height = 4}
for (d in demes$deme) {
  cat("\n## Inferred case counts v2- ", d, "\n")
  grid.arrange(trajs[[d]], tableGrob(reported %>% filter(deme == d) %>% t, theme = ttheme_minimal()), ncol = 2, widths = c(2,1))
  cat("\n")
}
```

## Reporting rate

```{r report-rate, fig.height = 4}
# 1.3. Reporting rate by day by case
reportrate <- gt_summary_v2 %>%
  filter(!type %in% c(china_deme, "OtherEuropean")) %>%
  left_join(case_data %>% rename(type = deme), by = c("type", "date"))  %>%
  replace_na(list(cases = 0)) %>%
  mutate(reportrate = ifelse(is.infinite(cases/Dmedian), NA, cases/Dmedian),
         rlow = ifelse(is.infinite(cases/Dhigh), NA, cases/Dhigh),
         rhigh = ifelse(is.infinite(cases/Dlow), NA, cases/Dlow)) %>%
  filter(!is.na(reportrate)) %>%
  # mutate(week = lubridate::week(date)) %>%
  # group_by(week, type) %>%
  # mutate(w_reportrate = mean(reportrate)) %>%
  ggplot() +
  geom_line(aes(x = date, y = reportrate, color = type), stat="identity") +
  geom_point(aes(x = date, y = reportrate, color = type), stat="identity") +
  geom_ribbon(aes(date, ymin = rlow, ymax = rhigh, fill = type), alpha = 0.1) +
  ylab("Reporting rate") +
  scale_color_manual(values = dcolors, name = "") +
  scale_fill_manual(values = dcolors, name = "") +
  scale_x_date(limits = c(ymd("2020-01-20"), ymd(MRS + 10)), date_breaks = "1 week", date_labels = "%b %d") +
  theme(#legend.position = "none",
        axis.title.x=element_blank()) +
  coord_cartesian(ylim = c(0,5))

reportrate
#
# ggexport(reported5days, filename = paste0(args$output_figure, "03.png"),
#          width = 1500, height = 1000, res = 300)

```


## Inferred case counts - Additional figures
```{r traj-sum, fig.height = 4}
# 1.1 Median and 95% CCI by deme in log scale
gribbon <- ggplot(gt_summary) +
  geom_ribbon(aes(date, ymin = Ilow, ymax = Ihigh, fill = type), alpha = 0.5) +
  geom_line(aes(date, Imedian, colour = type)) +
  ylab("Population size") +
  xlab("Date") +
  #labs(title = "SARS-CoV-2 early epidemics in Europe and Hubei",
       #subtitle = "Case trajectories Median and 95% CCI by deme") +
  scale_y_log10(labels = trans_format("log10", math_format(10^.x))) +
  scale_x_date(limits = c(ymd("2019-10-01"), ymd("2020-03-08")), date_breaks = "1 month", date_labels = "%b %d") +
  scale_fill_manual(name = "", values = dcolors) +
  scale_color_manual(name = "", values = dcolors) +
  theme(legend.position = c(0.2, 0.7),
        legend.box="horizontal",
        axis.title.x=element_blank())

gribbon

#
# ggexport(gribbon,
#          filename = paste0(args$output_figure, "01.png"),
#          width = 1500, height = 1000, res = 300)
```

<!-- ## First introduction -->
<!-- ```{r} -->
<!-- # 2. Events  -->
<!-- # 2.1. Time interval of first case (introduction) and first reported case to ECDC. -->

<!-- first_ecdc <- case_data %>%  -->
<!--   group_by(deme) %>%  -->
<!--   filter(cumcases != 0) %>%  -->
<!--   arrange(date) %>%  -->
<!--   slice(1)  %>%  -->
<!--   rename(dest = deme, min_date = date) -->

<!-- firstcase <- events %>% -->
<!--   filter(event == "M" | event == "O") %>% -->
<!--   mutate(dest = ifelse(is.na(dest), china_deme, dest)) %>% -->
<!--   group_by(dest, traj) %>% -->
<!--   arrange(time) %>% -->
<!--   slice(1) %>% -->
<!--   ungroup %>% -->
<!--   # Add row for Europe (all countries) -->
<!--   bind_rows(events %>% -->
<!--               filter(event == "M") %>% -->
<!--               group_by(traj) %>% -->
<!--               arrange(time) %>% -->
<!--               slice(1)  %>% -->
<!--               mutate(dest = "Europe")) -->

<!-- dates_firstcase <-  firstcase %>% -->
<!--   group_by(dest, min_date) %>% -->
<!--   summarize(Dmedian = median(date),  -->
<!--             Dlow    = quantile(date, 0.025, type = 1),  -->
<!--             Dhigh   = quantile(date, 0.975, type = 1), -->
<!--             .groups = "drop") %>% -->
<!--   mutate(difmedian = min_date - Dmedian, -->
<!--          diflow = min_date - Dlow, -->
<!--          difhigh = min_date - Dhigh)  -->

<!-- # 2.2. Time interval of first bith event (within region transmission) and first reported case -->
<!-- # to ECDC. -->
<!-- # How well did the countries detecting the first cases, there was already within region transmision? -->

<!-- # First introduction: Measure expected day of introduction. -->
<!-- firstbirth <- events %>% -->
<!--   filter(event == "B" | event == "O") %>% -->
<!--   group_by(dest, traj) %>% -->
<!--   arrange(time) %>% -->
<!--   slice(1) %>% -->
<!--   ungroup %>% -->
<!--   left_join(case_data %>% group_by(deme) %>% filter(cumcases != 0) %>% arrange(date) %>%slice(1)  %>% rename(dest = deme, min_date = date), by = "dest") %>% -->
<!--   bind_rows(events %>% -->
<!--               filter(dest != china_deme, event == "B") %>% -->
<!--               group_by(traj) %>% -->
<!--               arrange(time) %>% -->
<!--               slice(1)  %>% -->
<!--               mutate(dest = "Europe")) -->



<!-- dates_firstbirth <- firstbirth %>% -->
<!--   group_by(dest, min_date) %>% -->
<!--   summarize(Dmedian = median(date),  -->
<!--             Dlow    = quantile(date, 0.025, type = 1),  -->
<!--             Dhigh   = quantile(date, 0.975, type = 1), -->
<!--             .groups = "drop") %>% -->
<!--   mutate(difmedian = min_date - Dmedian, -->
<!--          diflow = min_date - Dlow, -->
<!--          difhigh = min_date - Dhigh) -->

<!-- p_firstbirth <- events %>% -->
<!--   filter(event == "B" | event == "O") %>% -->
<!--   bind_rows(events %>% -->
<!--               filter(dest != china_deme, event == "B") %>% -->
<!--               group_by(traj) %>% -->
<!--               arrange(time) %>% -->
<!--               slice(1)  %>% -->
<!--               mutate(dest = "Europe")) %>% -->
<!--   group_by(dest, traj) %>% -->
<!--   arrange(time) %>% -->
<!--   slice(1) %>% -->
<!--   ungroup %>% -->
<!--   left_join(case_data %>% group_by(deme) %>% filter(cumcases != 0) %>%  -->
<!--               arrange(date) %>% slice(1) %>% rename(dest = deme, min_date = date), by = "dest") %>% -->
<!--   replace_na(list(min_date = ymd("2020-01-25"))) %>% -->
<!--   filter(date <= min_date) %>% -->
<!--   count(dest) %>% -->
<!--   mutate(p = n/args$n * 100) -->

<!-- r_firstcases <- dates_firstcase%>% -->
<!--   left_join(dates_firstbirth, by = c("dest", "min_date")) %>% -->
<!--   left_join(p_firstbirth, by = "dest") %>% -->
<!--   replace_na(list(min_date = dates_firstcase$min_date[dates_firstcase$dest == "France"])) %>% -->
<!--   mutate(Country = dest, -->
<!--          `First Reported Case ECDC` = min_date, -->
<!--          `Median date (95% CCI) first introduction` = paste0(Dmedian.x, " (", Dlow.x, ", ", Dhigh.x, ")"), -->
<!--          `Median start date (95% CCI) within region transmission` = paste0(Dmedian.y, " (", Dlow.y, ", ", Dhigh.y, ")"), -->
<!--          `Percentage of trajectories within region transmission when first reported case` = paste0(p, "%")) %>% -->
<!--   select(c("Country", "First Reported Case ECDC", "Median date (95% CCI) first introduction", "Median start date (95% CCI) within region transmission", -->
<!--            "Percentage of trajectories within region transmission when first reported case")) -->


<!-- # 2.1. Time interval of first outcoming migration from a country compared to first reported case -->
<!-- # In which percentage of trajectories, transmission from that country would have been avoided if -->
<!-- # any ways of outcoming migrations were closed the day of first reported case. -->

<!-- firstmig <- events %>% -->
<!--   filter(event == "M") %>% -->
<!--   group_by(src, traj) %>% -->
<!--   arrange(time) %>% -->
<!--   slice(1) %>% -->
<!--   ungroup %>% -->
<!--   left_join(case_data %>% group_by(deme) %>% filter(cumcases != 0) %>% arrange(date) %>% slice(1)  %>% rename(src = deme, min_date = date), by = "src")  -->

<!-- dates_firstmig <- firstmig%>% -->
<!--   group_by(src, min_date) %>% -->
<!--   summarize(Dmedian = median(date),  -->
<!--             Dlow    = quantile(date, 0.025, type = 1),  -->
<!--             Dhigh   = quantile(date, 0.975, type = 1), -->
<!--             .groups = "drop") %>% -->
<!--   mutate(difmedian = min_date - Dmedian, -->
<!--          diflow = min_date - Dlow, -->
<!--          difhigh = min_date - Dhigh)  -->

<!-- p_firstmig <- events %>% -->
<!--   filter(event == "M") %>% -->
<!--   group_by(src, traj) %>% -->
<!--   arrange(time) %>% -->
<!--   slice(1) %>% -->
<!--   ungroup %>% -->
<!--   left_join(case_data %>% group_by(deme) %>% filter(cumcases != 0) %>% arrange(date) %>% slice(1)  %>% rename(src = deme, min_date = date), by = "src") %>% -->
<!--   filter(date >= min_date) %>% -->
<!--   count(src) %>% -->
<!--   mutate(p = n/args$n * 100) -->


<!-- r_firstmig <- dates_firstmig %>% -->
<!--   left_join(p_firstmig, by = "src") %>% -->
<!--   mutate(Country = src, -->
<!--          `First Reported Case ECDC` = min_date, -->
<!--          `Median date (95% CCI) first outgoing migration` = paste0(Dmedian, " (", Dlow, ", ", Dhigh, ")"), -->
<!--          `Percentage of trajectories first migration after first reported case` = paste0(p, "%")) %>% -->
<!--   select(-one_of(c(colnames(dates_firstmig), colnames(p_firstmig)))) -->


<!-- # Measure the time between the first migration into a country and the first birth in the country.  -->
<!-- # Or the first outcoming mirgration from the country and first birth.  -->
<!-- # This could be interesting to say if we should focus or not the screening and testing  -->
<!-- # capacities to detect incoming migrations or a more general strategy to find cases in  -->
<!-- # the population. Is it different for each country? -->

<!-- firsts <- firstcase %>% -->
<!--   select(dest, date, min_date) %>% -->
<!--   rename(deme = dest, introduction = date) %>% -->
<!--   left_join(firstbirth %>% select(date, dest) %>% rename(deme = dest, birth = date), by = "deme") %>% -->
<!--   left_join(firstmig %>% select(date, src) %>% rename(deme = src, mig = date), by = "deme") %>% -->
<!--   filter(deme != "Europe") %>% -->
<!--   mutate(dif1 = birth - introduction, -->
<!--          dif2 = mig - introduction) %>% -->
<!--   group_by(deme) %>% -->
<!--   summarize(dif1median = median(dif1),  -->
<!--             dif1low    = quantile(dif1, 0.025, type = 1),  -->
<!--             dif1high   = quantile(dif1, 0.975, type = 1), -->
<!--             dif2median = median(dif2),  -->
<!--             dif2low    = quantile(dif2, 0.025, type = 1),  -->
<!--             dif2high   = quantile(dif2, 0.975, type = 1), -->
<!--             .groups = "drop")  -->


<!-- r_firsts <- firsts %>% -->
<!--   mutate(Country = deme, -->
<!--          `Days between introduction and first within region transmision` = paste0(dif1median, " (", dif1low, ", ", dif1high, ")"), -->
<!--          `Days between introduction and first outgoing migration` = paste0(dif2median, " (", dif2low, ", ", dif2high, ")")) %>% -->
<!--   select(-one_of(colnames(firsts))) %>% -->
<!--   left_join(r_firstcases, by = "Country") %>% -->
<!--   left_join(r_firstmig, by = c("Country", "First Reported Case ECDC")) %>% -->
<!--   select(1,4,5,6,8,2,3,7,9) -->

<!-- write_tsv(r_firsts, paste0(args$output_table, "02.tsv")) -->

<!-- ``` -->

